Kindly Bent to Free Us

### 2019-07-11 kindly/soa.tex
checkpoint
cleanup for pre-submission

\section{Related work}

Haller, P., Odersky, M.: Capabilities for Uniqueness and Borrowing
\cite{DBLP:conf/ecoop/HallerO10}

Sing\#

John Tang Boyland and William Retert. Connecting effects and
uniqueness with adoption. In POPL, pages 283–295. ACM, 2005.
\cite{DBLP:conf/popl/BoylandR05}

@@

Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18} aims
to retrofit Haskell with linear types.
Linear Haskell introduces a new notion
of ``linear arrows'', noted $\multimap$ similar to linear logic,
which are functions who use their argument exactly once.
% By contrast, in \lang linearity is only decided
% by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
This design is easy to retrofit on top of an existing compiler
such as GHC, but has several disadvantages\footnote{
  See the in-depth discussion attached to the GHC proposal starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
Most relevant to \lang:
\begin{itemize}[leftmargin=*]
\item Their system does not admit subtyping for arrows and leverages
  $\eta$-expansion to pass unrestricted functions in linear
  contexts. This works in a non-strict language such as
  Haskell but can not be used in a strict setting.
\item
  While the GHC proposal implements some type inference,
  it is neither formalized nor complete.
\item
  Linear Haskell promotes a continuation-passing style with functions such as 
  \lstinline/withFile : path -> (file ->. file) ->. unit/
  to ensure that files are only used linearly.
  \lang follows System F\degree, Quill and Alms and
  adopts a direct style, where types themselves are
  described as affine or linear.
  We believe this leads to a more natural programming style
  and ease modular reasoning about linearity.
  In particular, using abstraction through modules,
  programmers only need to consider the module
  implementation to ensure that linear
  resources are properly handled.
% \item
%   Linear Haskell introduces a notion of linear monad to express
%   imperative code conveniently. Again, this solution is suitable in Haskell,
%   but less appropriate in other contexts. Our borrow system allow
%   to provide a more free-form programming style for imperative code.
% \item
%   Due to the previous remark, ensuring that a given object is never
%   aliased is fairly difficult in Linear Haskell, as one would need to ensure
%   that no non-linear function can manipulate it. On the other
\end{itemize}

Mezzo~\citep{DBLP:phd/hal/Protzenko14} is an ML-like language
with an extremely rich capability system which allows it to encode numerous
properties akin to separation logic~\citep{DBLP:conf/lics/Reynolds02}.
The main contribution of Mezzo is to push the boundary of what is possible
to do with a type system. This makes it far more expressive than \lang, but
also quite harder to use, and precludes any form of complete type
inference.

@@

Linear and affine discipline have also been used in non-functional
settings, notably in the context of low-level imperative programming
and object systems. In particular, \lang uses borrows and regions
which were initially developed in objects or
imperatives settings~\citep{DBLP:conf/popl/BoylandR05,DBLP:conf/pldi/GrossmanMJHWC02}.

@@

\subsection{Type-system features}

\lang relies on qualified types~\cite{DBLP:journals/scp/Jones94}
to introduce the necessary kind equalities for linear types. In
particular, we use HM(X)~\citep{DBLP:journals/tapos/OderskySW99} which
allow to use qualified type in an ML-like language with complete
type inference.
HM(X) has been shown to be compatible with subtyping,
bounded universal and existentials~\citep{DBLP:conf/icfp/Simonet03}
GADTs~\citep{DBLP:journals/toplas/SimonetP07},
and syntactic proof of soundness~\citep{DBLP:journals/entcs/SkalkaP02}.
This makes us confident that the system developed in \lang
could be applied to larger and more complex languages such as OCaml.

>>>>

\section{Related work}
\label{sec:related-work}

% Haller, P., Odersky, M.: Capabilities for Uniqueness and Borrowing
% \cite{DBLP:conf/ecoop/HallerO10}

% Sing\#

% John Tang Boyland and William Retert. Connecting effects and
% uniqueness with adoption. In POPL, pages 283–295. ACM, 2005.
% \cite{DBLP:conf/popl/BoylandR05}

@@

The goal of Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
(LH) is to retrofit linear types to Haskell.
LH introduces a new notion
of ``linear arrows'', written $\multimap$ similar to linear logic,
which are functions that \emph{use} their argument exactly once.
% By contrast, in \lang linearity is only decided
% by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
This design is easy to retrofit on top of an existing compiler
such as GHC, but there are quite controversial discussions about it\footnote{
  See the in-depth discussion attached to the GHC proposal starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
Most relevant to \lang:
\begin{itemize}[leftmargin=*]
\item LH does not admit subtyping for arrows and requires
  $\eta$-expansion to pass unrestricted functions in linear
  contexts. This approach is acceptable in a non-strict language such as
  Haskell but changes the semantics in a strict setting.
\item
  The GHC implementation of LH performs type inference, but
  it is neither formalized nor complete.
\item
  LH promotes a continuation-passing style with functions such as 
  \lstinline/withFile : path -> (file ->. file) ->. unit/
  to ensure linear use of resources. This style leads to problems with
  placing the annotation on, e.g., the IO monad.
  \lang follows System F\degree, Quill, and Alms, all of which support
  resource handling in direct style, where types themselves are
  described as affine or linear. (Or course, continuation-passing
  style is also supported.)
  We expect that the direct approach eases modular reasoning about linearity.
  In particular, using abstraction through modules,
  programmers only need to consider the module
  implementation to ensure that linear
  resources are properly handled.
% \item
%   Linear Haskell introduces a notion of linear monad to express
%   imperative code conveniently. Again, this solution is suitable in Haskell,
%   but less appropriate in other contexts. Our borrow system allow
%   to provide a more free-form programming style for imperative code.
% \item
%   Due to the previous remark, ensuring that a given object is never
%   aliased is fairly difficult in Linear Haskell, as one would need to ensure
%   that no non-linear function can manipulate it. On the other
\end{itemize}

Mezzo~\citep{DBLP:phd/hal/Protzenko14} is an ML-like language
with an extremely rich capability system which is able to encode numerous
properties akin to separation logic~\citep{DBLP:conf/lics/Reynolds02}.
Mezzo explores the  boundaries of the design space of type systems for
resources. Hence, it is more expressive than \lang, but
much harder to use. The Mezzo typechecker relies on explicit
annotations and it is not known whether type inference for Mezzo is possible.

@@

Linear and affine discipline have also been used in non-functional
settings, notably in the context of low-level imperative programming
and object systems. In particular, \lang uses borrows and regions
which were initially developed in the context of imperative and
object-oriented
programming~\citep{DBLP:conf/popl/BoylandR05,DBLP:conf/pldi/GrossmanMJHWC02}.

@@

\subsection{Type-system features}
%
\lang relies on constrained types
to introduce the kind inequalities required for linear types. In
particular, we use HM(X)~\citep{DBLP:journals/tapos/OderskySW99} which
allow to use constrained types in an ML-like language with complete
type inference.
HM(X) has been shown to be compatible with subtyping,
bounded universal and existentials~\citep{DBLP:conf/icfp/Simonet03},
GADTs~\citep{DBLP:journals/toplas/SimonetP07},
and there exists a syntactic soundness proof~\citep{DBLP:journals/entcs/SkalkaP02}.
These results make us confident that the system developed in \lang
could be applied to larger and more complex languages such as OCaml.

Alternatively, we could have based the system on qualified
types~\cite{DBLP:journals/scp/Jones94}, similarly to Quill. This
choice would also be sustainable as qualified types is part of the
foundation of Haskell's type system.
